Schema Change: Challenge Problems

### 2024-05-03 paper.tex
Start of Related Work

@@

In a SQL database, \texttt{Customers.Name} would be a primary key and \texttt{Orders.Customer} would be a foreign key. But before \texttt{Customers.Name} can be made a primary key the two occurrences of \texttt{Wile E Coyote} must be deduplicated one way or another~\cite{dedupe}. Given the dominance of SQL databases it is not surprising that there are many tools for performing schema change on them~\cite{RailsMigrations, liquibase}. We do find it surprising how low-level these tools are. There is support for the basic SQL DDL commands like adding and dropping tables, columns, or constraints, but no higher level operations that would help solve our problem. One is left to write custom SQL, which can become difficult if the migration needs to be performed online without bring down the database. There is a rich theory of relational database \textit{normalization} based on \textit{functional dependencies}~\cite{Molina08} but that theory has not been implemented into schema change operations in databases -- all you get is SQL. \citet{ambler06} define higher-level patterns of schema change like \textit{Split Table} that sketch SQL implementations but don't provide everything required for Extract Entity. These difficulties fairly beg for new linguistic abstractions, though programming patterns and APIs are reasonable pragmatic solutions to our challenge.

@@

\paragraph{Applications.}
One of the earliest systems to combine documents with computational elements
is Boxer \cite{diSessa86}. The challenges in this section apply to a wide range of
systems that can be seen as its successors. The specific scenario discussed here is
inspired by the use of Notion~\cite{Notion24}, which is a collaborative structured document
editing system. Treating documents as end-user programs has recently
been explored in Potluck~\cite{Litt2023}, which is local-only but includes formulas similar
to those in our challenge.

>>>>

\section{Related Work}

Schema change and data migration are a widespread problem in practice for SQL databases, attested by the many tools that have been created to help. Some tools, like Rails Migrations~\cite{RailsMigrations} provide a Domain Specific Language (DSL) for specifying and executing schema changes. Other tools, such as those in Liquibase~\cite{liquibase} or PlanetScale\cite{planetscale} are based on textually comparing SQL Data Definition Language (DDL) statements that define the database schema. The differences between these DDL schema can be used to infer a delta of DDL statements to change one into the other. Unfortunately these inferences must sometimes guess at the user's intent: for example has a column been renamed or has it been deleted and a new one created? The difference matters a lot for the data in that column! EdgeDB~\cite{edgedb} asks the developer to resolve such ambiguities with a series of questions.

All of these tools end up emitting SQL DDL statements to change the schema. Unfortunately that does not migrate the data. Some databases add limited features for migration. For example SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes. MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for schema migration. A developer must write custom SQL code to transform data from the old to the new schema. This custom migration code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards. In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema change patterns, including typical strategies for online migration and example SQL code.

There is surprisingly little discussion of schema migration in the database literature nor support from the major database vendors. There is indeed a rich theory of database normalization~\cite{Molina08} but that is presumed to happen before the database is populated with data. To a large extent this entire problem area has been relegated to application developers and `dev ops'. Yet prima facia, schema migration can only be done throughly with direct support inside the database. EdgeDB~cite{edgedb} stands out for its improved support of schema migration through a proprietary language of data transformations but is still limited by the capabilties of its underlying PostgresSQL database.

\begin{comment}

Newly discovered thread of DB research on schema change:
https://en.wikipedia.org/wiki/Schema_evolution
https://www.researchgate.net/publication/220710372_Schema_Evolution_in_Wikipedia_-_Toward_a_Web_Information_System_Benchmark
https://dl.acm.org/doi/10.14778/1453856.1453939
https://dl.acm.org/doi/10.1145/3448016.3452842
https://dl.acm.org/doi/10.14778/3565816.3565818
https://link.springer.com/chapter/10.1007/978-3-030-58135-0_10

Schema change IDEs
https://www.red-gate.com/solutions/overview
https://www.jetbrains.com/datagrip/features/

Smalltalk images
	depends on dynamic types and adding new fields to objects
	Gemstone added complex API

“unstructured data”: JSON and RDF used with untyped PL like Javascript or Python
Like Smalltallk, small set of changes can be handled gradecfully, but in general need to write custom code

dataframe diff
https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf

Tijs on Live programming and model-driven engineering
https://ir.cwi.nl/pub/23770
https://dl.acm.org/doi/10.1007/s10270-017-0608-7
https://dl.acm.org/doi/10.1145/3276604.3276611
https://ir.cwi.nl/pub/21858

Geoffreu on Local-first
https://www.inkandswitch.com/cambria/
https://riffle.systems/papoc22.pdf

Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change

Version control for dbs
https://people.eecs.berkeley.edu/~brewer/cs262b/update-conflicts.pdf
https://github.com/dolthub/dolt
https://github.com/dolthub/doltgresql
https://irmin.org
https://doc.replicache.dev/concepts/how-it-works

\end{comment}

@@

In a SQL database, \texttt{Customers.Name} would be a primary key and \texttt{Orders.Customer} would be a foreign key. But before \texttt{Customers.Name} can be made a primary key the two occurrences of \texttt{Wile E Coyote} must be deduplicated one way or another~\cite{dedupe}.

\textcolor{red}{this paragraph is now redundant with Related Owrk}
Given the dominance of SQL databases it is not surprising that there are many tools for performing schema change on them~\cite{RailsMigrations, liquibase}. We do find it surprising how low-level these tools are. There is support for the basic SQL DDL commands like adding and dropping tables, columns, or constraints, but no higher level operations that would help solve our problem. One is left to write custom SQL, which can become difficult if the migration needs to be performed online without bring down the database. There is a rich theory of relational database \textit{normalization} based on \textit{functional dependencies}~\cite{Molina08} but that theory has not been implemented into schema change operations in databases -- all you get is SQL. \citet{ambler06} define higher-level patterns of schema change like \textit{Split Table} that sketch SQL implementations but don't provide everything required for Extract Entity. These difficulties fairly beg for new linguistic abstractions, though programming patterns and APIs are reasonable pragmatic solutions to our challenge.

@@

\paragraph{Applications.}
One of the earliest systems to combine documents with computational elements
is Boxer \cite{diSessa86}. The challenges in this section apply to a wide range of
systems that can be seen as its successors. The specific scenario discussed here is
inspired by the use of Notion~\cite{notion}, which is a collaborative structured document
editing system. Treating documents as end-user programs has recently
been explored in Potluck~\cite{Litt2023}, which is local-only but includes formulas similar
to those in our challenge.
